---
title: "MovieLens Recommendation Systen"
author: "Fernando Scalice Luna"
date: "5/31/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r include=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data

set.seed(1) # if using R 3.6.0: set.seed(1, sample.kind = "Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

```{r include=FALSE}
# Required libs to run this project
library(stringr)
library(gridExtra)
library(gtable)
library(grid)
```

# 1. Executive Summary

Based on a large data set of movie ratings, a machine learning algorithm was developed to predict user ratings.

This data set contains 10,000,054 ratings and 95,580 tags applied to 10,681 movies by 71,567 users of the online movie recommender service MovieLens.

The metric we're going to use to evaluate how good our model is will be the RMSE.

4 models were developed considering:

1. Simple average of ratings
2. Movie Effects
3. User Effects
4. Movie Age Effects

The final model resulting in the lowest RMSE (~0.865) is a combination of the first three models.

We finally tuned our model to include appropriate regularization and validated against our test set. The final result was a RMSE of 0.825.

# 2. Introduction - Data Exploration

After downloading the files, creating the train and test set, we can start exploring the data set creating some charts to better understand the problem.

Let's start with a distribution of ratings
```{r}
edx %>% group_by(movieId) %>%
  summarise(n = n()) %>%
  ggplot(aes(n)) + 
  geom_histogram(fill = "darkred", color = "white", bins = 30) +
  scale_x_log10() +
  ggtitle("Movies Ratings") +
  xlab("Number of ratings (log10)") +
  ylab("Number of movies")
```

Now let's see the distribution of ratings by users
```{r}
edx %>% group_by(userId) %>%
  summarise(n = n()) %>%
  ggplot(aes(n)) + 
  geom_histogram(fill = "darkred", color = "white", bins = 30) +
  scale_x_log10() +
  ggtitle("User Ratings") +
  xlab("Number of ratings (log10)") +
  ylab("Number of users")
```

```{r}
edx %>% group_by(userId) %>%
  summarise(b_u = mean(rating)) %>%
  ggplot(aes(b_u)) + 
  geom_histogram(fill = "darkred", color = "white", bins = 30) +
  ggtitle("Ratings by User") +
  xlab("Rating") +
  ylab("Number of ratings")

mean(edx$rating)
```
Ratings tend to fall in the range of 3 to 4.5.

## Exploring the Age of the movies
First we need to extract the Release Date from the movie title and then calculate the age.
```{r}
release_pattern <- "(\\()(\\d{4})(\\))"
edx_age <- edx %>% mutate(release = as.numeric(str_match(edx$title, release_pattern)[,3]))

# Checking if extraction is ok
identical(as.numeric(str_match(edx_age$title, release_pattern)[,3]), edx_age$release)

# Calculating the AGE
edx_age <- edx_age %>% mutate(age = 2019 - release)
```

Now we can see the distribution of rating by age
```{r}
edx_age %>% 
  ggplot(aes((age))) +
  geom_histogram(fill = "darkred", color = "white", bins = 30) +
  ggtitle("Ratings by Age") +
  xlab("Age") +
  ylab("Number of ratings")

  mean(edx_age$age)
```
It seems movies from the 90's tend to get more ratings. Let's dig a little deeper to see why.

```{r}
# Number of movies by Age
edx_age %>% group_by(movieId) %>% summarise(n = n(), age = first(age)) %>%
    ggplot(aes((age))) +
    geom_histogram(fill = "darkred", color = "white", bins = 30) +
    ggtitle("Movies by Age") +
    xlab("Age") +
    ylab("Number of movies")

```
We see from the plot above that there are more movies from the 90's!

Now, let's check if age might influence rating
```{r}
avg_rating_by_age <- edx_age %>% group_by(age) %>% summarise(avg_rating = mean(rating))

avg_rating_by_age %>%
  ggplot(aes(x = age, y = avg_rating)) +
  geom_point() +
  geom_smooth() +
  ggtitle("Average rating by Movie Age") +
  xlab("Movie Age") +
  ylab("Average Age")
```
It seems movie age plays a role in estimating the rate of a movie.

# 3. Building the Model

Before we start building our model, let's define our RMSE function.

```{r}
RMSE <- function(true_ratings, predicted_ratings){ 
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

## 3.1 First Model - Simple Average
Our first model is a simple average of ratings.

```{r}
mu_hat <- mean(edx$rating) 
mu_hat

naive_rmse <- RMSE(validation$rating, mu_hat) 
naive_rmse

rmse_results <- tibble(method = "Just the average", RMSE = naive_rmse)
```
As we can see the RMSE is above 1. We need to improve it.

## 3.2 Second Model - Movie Effect
In our second model we will include the movie effect to see if we can lower the RMSE
```{r}
mu <- mean(edx$rating) 
movie_avgs <- edx %>%
  group_by(movieId) %>% summarize(b_i = mean(rating - mu))

movie_avgs %>% ggplot(aes(b_i)) +
  geom_histogram(fill = "darkred", color = "white", bins = 30)

predicted_ratings <- mu + validation %>% 
  left_join(movie_avgs, by='movieId') %>% 
  pull(b_i)

movie_rmse <- RMSE(validation$rating, predicted_ratings) 
rmse_results <- bind_rows(rmse_results,
                          tibble(method="Movie Effect Model", RMSE = movie_rmse))
rmse_results

```
This model improved our predictions and our RMSE is now at 0.943. However, it's not enough to pass.

## 3.3 Third Model - User Effect
Our third model will include the user effect.
```{r}
user_avgs <- edx %>% left_join(movie_avgs, by='movieId') %>% 
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))

predicted_ratings <- validation %>% 
  left_join(movie_avgs, by='movieId') %>% 
  left_join(user_avgs, by='userId') %>% 
  mutate(pred = mu + b_i + b_u) %>% 
  pull(pred)

user_rmse <- RMSE(validation$rating, predicted_ratings)
rmse_results <- bind_rows(rmse_results, 
                          tibble(method="Movie + User Effects Model", RMSE = user_rmse))

rmse_results
```
We're getting better. Our model now has a RMSE of 0.865. Although it's enough to pass, let's see if we can improve even more.

## 3.4 Fourth Model - Age Effect
Our fourth model will then include the age effect. 
As we saw that rating varies with age, we will use a smooth function using loess to fit the data.
```{r}
total_span <- diff(range(avg_rating_by_age$age))
span <- 30/total_span

avg_rating_by_age.loess <- loess(avg_rating ~ age, 
                                 degree = 1, 
                                 span = span, 
                                 data = avg_rating_by_age)

avg_rating_by_age %>% mutate(smooth = avg_rating_by_age.loess$fitted) %>% 
  ggplot(aes(age, avg_rating)) +
  geom_point(size = 3, alpha = .5, color = "black") + 
  geom_line(aes(age, smooth), size = 1, color="red")
```

In order to test it, we need to make a little adaptation to our validaton set to include the age
```{r}
age_avgs <- edx_age %>% 
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  group_by(age) %>%
  summarize(b_a = mean(predict(avg_rating_by_age.loess, newdata = age) - mu - b_i - b_u))

validation <- validation %>% 
  mutate(release = as.numeric(str_match(title, release_pattern)[,3])) %>% 
  mutate(age = 2019 - release)

predicted_ratings <- validation %>% 
  left_join(movie_avgs, by='movieId') %>% 
  left_join(user_avgs, by='userId') %>%
  left_join(age_avgs, by='age') %>%
  mutate(pred = mu + b_i + b_u + b_a) %>% 
  pull(pred)

age_rmse <- RMSE(validation$rating, predicted_ratings)
rmse_results <- bind_rows(rmse_results, 
                          tibble(method = "Movie, User and Age Effects Model", RMSE = age_rmse))
rmse_results
```
As we see, age does not improve our model, so let's keep it out from our model.

# 4. Results

Now we have our model almost ready. It's time to tune it and penalize large estimates from small sample sizes.  
```{r}
lambdas <- seq(0, 5, .25)
rmses <- sapply(lambdas, function(l){
  mu <- mean(edx_age$rating)
  
  b_i <- edx_age %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx_age %>%
    left_join(b_i, by="movieId") %>% 
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- edx_age %>%
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>% .$pred
  
  return(RMSE(predicted_ratings, edx_age$rating))
})
```

```{r}
qplot(lambdas, rmses)
lambdas[which.min(rmses)]
```
The best lambda for our model is 0.5.

```{r}
rmse_results <- bind_rows(rmse_results,
                          tibble(method="Regularized Movie + User Effect Model",
                                 RMSE = min(rmses)))

rmse_results %>% knitr::kable()
```

Now we have a final model with a RMSE of 0.856

We are not ready yet. Let's finally test our model against the validation set and see the results

# 5. Conclusion

Testing our model against the validation is the final step to make sure we have built a decent model.
```{r}
mu <- mean(validation$rating)
l <- lambdas[which.min(rmses)]

b_i <- validation %>%
  group_by(movieId) %>%
  summarize(b_i = sum(rating - mu)/(n()+l))

b_u <- validation %>%
  left_join(b_i, by="movieId") %>% 
  group_by(userId) %>%
  summarize(b_u = sum(rating - b_i - mu)/(n()+l))

predicted_ratings <- validation %>%
  left_join(b_i, by = "movieId") %>%
  left_join(b_u, by = "userId") %>%
  mutate(pred = mu + b_i + b_u) %>% .$pred

RMSE(predicted_ratings, validation$rating)
```
As we can see, our model is able to predict the ratings with a RMSE of 0.825.

