---
title: "MovieLens Recommendation Systen"
author: "Fernando Scalice Luna"
date: "5/31/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Executive Summary

This data set contains 10000054 ratings and 95580 tags applied to 10681 movies by 71567 users of the online movie recommender service MovieLens.

Users were selected at random for inclusion. All users selected had rated at least 20 movies. Unlike previous MovieLens data sets, no demographic information is included. Each user is represented by an id, and no other information is provided.

The data are contained in three files, movies.dat, ratings.dat and tags.dat. Also included are scripts for generating subsets of the data to support five-fold cross-validation of rating predictions. More details about the contents and use of all these files follows.

RMSE


# Introduction - Data Exploration

```{r}
edx %>% group_by(movieId) %>%
  summarise(n = n()) %>%
  ggplot(aes(n)) + 
  geom_histogram(fill = "darkred", color = "white", bins = 30) +
  scale_x_log10() +
  title("Movies Ratings")

edx %>% group_by(userId) %>%
  summarise(n = n()) %>%
  ggplot(aes(n)) + 
  geom_histogram(fill = "darkred", color = "white", bins = 30) +
  scale_x_log10() +
  title("User Ratings")
```



## Methods


## Results


## Conclusion


